Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-08-01 analysis/paper/paper.Rmd
Discussion

# Introduction
Variation in population patterns are regarded important drivers and consequences of cultural variation, and of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to constitute a 'site'. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2020; @crema2022 for a review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null-model. These null-models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null-model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a total of ^14^C-dates. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, further reducing the sample of ^14^C-dates to. Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R package *ADMUR* [@timpson2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* [@crema2021] for Monte Carlo simulation and model comparison.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product for each shoreline date against the probability of the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probability and is not, critically [@timpson2021, 3], found by using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected. As is also indicated by the BIC values, this model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development. Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. X--9000 BCE, followed by a significant positive peak just after 9000 BCE. The process of decline is then interrupted by positive deviations at, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation of more magnitude than that expected under the exponential model between X and X BCE. Finally, a positive deviation occurs again at X--5000 BCE. While all of the above deviations should also be considered with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that are disregarded here.

@@

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the standard models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8600 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022]. While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind the observed pattern.

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2008; @bjerck2017]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period following the deglaciation could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak starts around 8800 BCE and concludes around 8000 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. This logistic growth in the RSPD has a corresponding positive deviation in the SSPD, which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@berg-hansen2022 argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. Their analysis covers a more limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. However, in line with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and with a smaller magnitude around 7100--6900 BCE.

Subsequently, the SSPD deviates negatively from the exponential decay from c. 6800 to c. 5800 BCE, a decrease that was also noted by @berghansen2022. This is followed by a positive deviation just before the 4th millennium BCE. The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway that occurs around the same time, which is argued to reach its peak around 6000 BCE. It is also worth noting that there appears to be a corresponding negative deviation from the logistic null model in the RSPD just before 6000 BCE. However, given that the model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD, given the present data.

>>>>

# Introduction
Variation in population patterns are regarded as important drivers and potential consequences of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population size and density [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009].

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the retreat of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at their time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years, or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. The point-based estimates do therefore not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a 'site' [@palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2021; @crema2022 for reviews].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to be older than around 9300 BCE in the region [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83]. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, leading to the final sample of 310 ^14^C-dates. Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at X. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for Monte Carlo simulation and model comparison.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product for each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probability and is not, critically [@timpson2021, 3], found by using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected. As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development. Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded here.

@@

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of a more forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic in the region [e.g. @lindblom1984; @fuglestvedt2018, 16--17], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

A common interpretation of the Early Mesolithic in Norway is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2017; @bjerck2008; @fuglestvedt2012]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE and concluding around 8000 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that has plateaued in the first centuries after 8000 BCE. The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

<!-- @berg-hansen2022 found an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. Their analysis covers a more limited temporal span and geographical area, and is based on the aggregation of point-based estimates of shoreline dates within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. However, consistent with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and, although with a smaller magnitude, around 7100--6900 BCE. -->
